---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = NULL
)
```

# khronos

<!-- badges: start -->
[![R-CMD-check](https://github.com/tesselle/khronos/workflows/R-CMD-check/badge.svg)](https://github.com/tesselle/khronos/actions)
[![codecov](https://codecov.io/gh/tesselle/khronos/branch/master/graph/badge.svg)](https://codecov.io/gh/tesselle/khronos)

[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->

## Overview

A toolkit for absolute dating and analysis of chronological patterns. This package includes functions for chronological modeling and dating of archaeological assemblages from count data.

**Initial development is in progress.**

## Installation

You can install the released version of **khronos** from [CRAN](https://CRAN.R-project.org) with:

```{r cran-installation, eval=FALSE}
install.packages("khronos")
```

And the development version from [GitHub](https://github.com/) with:

```{r gh-installation, eval=FALSE}
# install.packages("remotes")
remotes::install_github("tesselle/khronos")
```

## Contributing

Please note that the **khronos** project is released with a [Contributor Code of Conduct](https://www.tesselle.org/conduct.html). By contributing to this project, you agree to abide by its terms.
